# Reproducible Research: Peer Assessment 1

```{r global_options, echo=FALSE}
opts_chunk$set(echo=FALSE)
```
## Loading and preprocessing the data
```{r}
# Prep download folder
if (!file.exists("data")) {
  dir.create("data")
}

# This was a zip file so let's extract the actual data files from it
if (!file.exists("./data/activity.csv")) {
  unzip("activity.zip", overwrite=FALSE, exdir="./data")
}

raw_data <- read.csv("./data/activity.csv", stringsAsFactors=FALSE, na.strings="NA")

str(raw_data)
summary(raw_data)
```

## What is mean total number of steps taken per day?
```{r}
# The following is going to be redundant ways to calcuate the sum of a column grouped by another column
# but I want to do both ways for posterity sake.
# (1) Use tapply
steps_per_day_1 <- tapply(raw_data$steps, raw_data$date, FUN=sum)
barplot(steps_per_day_1, main="Steps per Day", xlab="Day", ylab="Total Steps")

# (2) Use aggregate
steps_per_day_2 = aggregate(raw_data$steps, by=list(raw_data$date), FUN=sum)
names(steps_per_day_2) <- c("date", "total_steps")

mn <- mean(steps_per_day_2$total_steps, na.rm=TRUE)
md <- median(steps_per_day_2$total_steps, na.rm=TRUE)
i <- which.max(steps_per_day_2$total_steps)
message(sprintf("The maximum number of steps, %s, occurred on %s", steps_per_day_2[i, "total_steps"], steps_per_day_2[i, "date"]))
message(sprintf("Mean: %s\nMedian: %s", round(mn), md))
# TODO Should I just replace these with a call to "summary". 
#summary(steps_per_day_2$total_steps)
  
```

## What is the average daily activity pattern?
```{r}
avg_steps_per_interval = aggregate(raw_data$steps, by=list(raw_data$interval), FUN=mean, na.rm=TRUE)
names(avg_steps_per_interval) <- c("interval", "avg_steps")
plot(avg_steps_per_interval$interval, avg_steps_per_interval$avg_steps, type="l", main="Average Steps per 5-min Interval", xlab="5-m Interval", ylab="Average Steps")


i <- which.max(avg_steps_per_interval$avg_steps)
interval_int <- avg_steps_per_interval[i, "interval"]
hour <- floor(interval_int / 100)
min <- interval_int %% 100
message(sprintf("The maximum average number of steps, %s, occurred on interval %s (i.e. %02d:%02d)", round(avg_steps_per_interval[i, "avg_steps"]), interval_int, hour, min))

```


## Imputing missing values
### My technique for imputing assumes that the missing value at a given interval will be replaced with the average value at that same interval.
```{r}
# Display the number of missing values
message(sprintf("The steps column has %s entries with missing data (i.e. NA values)", 
                sum(is.na(raw_data$steps) == TRUE)))
# Or I could do this with summary(raw_data$steps)[7]


# I will assume the person behaves roughly the same each day. This is probably not 
# the case for weekends and weekdays but my assumption will remain.
# So whenever I find a missing value, I will impute that value with the average 
# value at the time from avg steps table.
head(avg_steps_per_interval)
imp_data <- raw_data
# The simplest way to do this is to build up a dataset for the same number of days as the original except everyday has the average 
# steps per 5-min interval. Then we just replace the NA value in the raw_data with the avg value at the same exact location. 
# I found it non-trivial to do an inner join with two tables that don't have the same length.
num_days = length(imp_data$steps) / length(avg_steps_per_interval$avg_steps)
avg_steps_per_interval <- do.call("rbind", replicate(num_days, avg_steps_per_interval, simplify=FALSE))
missing <- is.na(imp_data$steps)
# Only overwrite the NA (i.e. missing) values
imp_data$steps[missing] <- avg_steps_per_interval$avg_steps[missing] 

# Display the histogram
imp_steps_per_day <- tapply(imp_data$steps, imp_data$date, FUN=sum)
barplot(imp_steps_per_day, main="Steps per Day", xlab="Day", ylab="Total Steps")

# Display the mean and median
mn <- mean(imp_steps_per_day)
md <- median(imp_steps_per_day)
message(sprintf("Mean: %s\nMedian: %s", round(mn), round(md)))

```

## Are there differences in activity patterns between weekdays and weekends?
```{r}
# Create new column representing if the day is a weekday or not.
# Create and new vector and initialize it to all "weekday"
workweek <- rep("weekday", length(imp_data$steps))
# now map the actual dates to days of the week and then determine which of those are Sat or Sun
days <- weekdays(as.Date(imp_data$date))
weekends <- days %in% c("Saturday", "Sunday")
workweek[weekends] <- "weekend"
# convert it to a factor and add this column to our data frame
workweek <- as.factor(workweek)
imp_data$workweek <- workweek

# Calculate the average steps during the workweek and during the weekend. 
avg_steps_per_interval_weekday = aggregate(imp_data$steps, by=list(imp_data$interval, imp_data$workweek), FUN=mean, na.rm=TRUE)
names(avg_steps_per_interval_weekday) <- c("interval", "workweek", "avg_steps")
# Separate out the data just for the workweek and just for the weekend
workweek_data<-avg_steps_per_interval_weekday[avg_steps_per_interval_weekday$workweek == "weekday",c("interval", "avg_steps")]
weekend_data<-avg_steps_per_interval_weekday[avg_steps_per_interval_weekday$workweek == "weekend",c("interval", "avg_steps")]
# Now create the panel plot
par(mfrow=c(2,1))
plot(workweek_data$interval, workweek_data$avg_steps, type="l", main="Avg Steps During Workweek", xlab="5-m Interval", ylab="Average Steps")
plot(weekend_data$interval, weekend_data$avg_steps, type="l", main="Avg Steps During Weekend", xlab="5-m Interval", ylab="Average Steps")


```


